Enabling Ambient Backscatter \\Using a Low-Cost Software Defined Radio

### 2017-04-07 paper.tex

\subsection{Receiver}
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{./fig/receiver_arch}
\caption{Receiver architecture from a system point of view. Signal flow is from left to right.}
\label{fig:receiver_arch}
\end{figure}
C++ has been used to implement the receiver. We provide the code
publicly available under
\url{https://github.com/s3xm3x/backscatterBASKReceiver}. 
%A highly
%sophisticated bus system has been developed to exchange data between the
%different components of the system. 
Figure
\ref{fig:receiver_arch} shows the architecture of our receiver, the
signal flows from the left to the right.
%\textit{Librtlsdr} (cf. \cite{steve-m_librtlsdr}), which is based on
%\textit{libusb}, is used to transfer the data from the TV stick into our
%program. As described in section~\ref{sub:rtl2832}, the RTL2832 mixes
%down the high frequency to a intermediate frequency (both values are
%software adjustable). Subsequently the data is mixed with sine and
%cosine, low-pass filtered and finally transferred to the digital domain.
The raw I/Q data is available as two 8-bit values (real and imaginary)
from the RTL-SDR. With
these two values, the phase and magnitude of the data can be
determined for every sample. The received signal can be represented as:
\begin{equation}
	I+j\,Q = abs(I,Q) \cdot e^{j \times ang(I,Q)} 
\end{equation}

@@

The resulting values are sent through the bus to the decoder. After the
decoder received the samples it decides when a frame is sent or when the
channel is idle.  
The decoder also includes a function to correlate the
received bitstream with an expected pattern. Hence it is able to
determine the bit error ratio (BER). 
Furthermore the code we have
written so far subsumes different simulators for e.g. playing back
recoreded data. Another handy utility, which has been written in Octave,
is a tool (an oscilloscope) to look at the data, which is written into
files by the demodulator. An example of a recorded data stream with 25
kS/s can be seen in figure
\ref{fig:transmission}.  \begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{./fig/transmission}
\caption{Start of a transmission of 101010.. with 1 kbit/s with idle state before. Data from the Octave oscilloscope. Amplitude is quite low with an average of 12 \% of the maximum, which is 127.5. This is due to a low signal strength.}
\label{fig:transmission}
\end{figure}

@@

We have used the textit{RTL-SDR} to observe the space variations of the
signal strength. \textit{Octave} scripts have been written to aid in
this purpose. These scripts are available online under
\url{https://github.com/s3xm3x/RTLSDRspecAn}. The scripts perform a
frequency sweep through the desired band. The obtained samples are then
transferred from the time domain to the frequency domain using a fast
fourier transformation (FFT).

Because the RTL-SDR has a maximum (stable) sampling rate of 2.4 MSPs, we
are limited to a maximum simultaneous acquisition band of 1.2 MHz. We
are keep a safety margin to the maximum capabilities and sample the
desired frequency band at 900 kHz intervals and stitch these bands
together to form the overall frequency sweep. This approach is valid
only because we are not interested in real-time data.  With this
approach we are able to scan a 20 MHz band in roughly two minutes.

During the scanning process the gain is set to 40.2 dB. To get the
signal strength we simply take the average over a 10 MHz band around the
center frequency of the desired TV signal. This approach can be
justified by the fact, that DVB-T is specified up to 10 MHz bandwidth.
And the local TV signal provider is to its own statements using DVB-T 2.
The signal \ensuremath{|H|} is calculated as follows.  
\begin{equation}
	|H| = \Biggl| \sum_{n=0}^N FFT\biggl( \Re\{ s_{\text{band,n}}(t) \} \biggr) \Biggr|
\end{equation}     
where \ensuremath{N} is the total number of intervals accumulated over
the entire frequency sweep. The FFT
is only carried out over the real values, which are received from the
\textit{RTL-SDR}. \ensuremath{s_{\text{band,n}}(t)} is the signal in the
time domain of one interval (one subband). There are as many samples taken in one band as
needed for an FFT of size 2048. To finally get \ensuremath{|H_{tv,avg}|} one has to normalize everything and calculate the power. Before doing that a simple smoothing algorithm (FIR) is applied on \ensuremath{|H|}.
\begin{equation}
	|H_{tv,avg}| = 20 \cdot log (|H|) - 20 \cdot log(|H_0|)
\end{equation}   
where \ensuremath{|H_0|} is a normalization value measured at a certain
point in space. 
\begin{figure}[h]
	\centering
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/haversine}
	\end{minipage}
	\hfill
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/heatmap_626mhz}
	\end{minipage}
	\caption{Left: Signal strength fading against distance relative
to the spot where the maximum signal strength has been captured (TV
towers). 474 MHz and 626 MHz belong to a tower in Uppsala, Vedyxa and
594 MHz belongs to a tower in Uppsala, Rickomberga. Distances have been
calculated with the Haversine formula. Right: Heatmap of the 626 MHz TV
signal in Uppsala.}

>>>>

\subsection{Receiver}
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{./fig/receiver_arch}
\caption{Receiver architecture from a system point of view. Signal flow is from left to right.}
\label{fig:receiver_arch}
\end{figure}
We provide our C++ receiver code publicly available under
\url{https://github.com/s3xm3x/backscatterBASKReceiver}
%A highly
%sophisticated bus system has been developed to exchange data between the
%different components of the system. 
Figure
\ref{fig:receiver_arch} shows the architecture of our receiver, the
signal flows from the left to the right.
%\textit{Librtlsdr} (cf. \cite{steve-m_librtlsdr}), which is based on
%\textit{libusb}, is used to transfer the data from the TV stick into our
%program. As described in section~\ref{sub:rtl2832}, the RTL2832 mixes
%down the high frequency to a intermediate frequency (both values are
%software adjustable). Subsequently the data is mixed with sine and
%cosine, low-pass filtered and finally transferred to the digital domain.
The raw I/Q data is available as two 8-bit values (real and imaginary)
from the RTL-SDR. With
these two values, the phase and magnitude of the data can be
determined for every sample. The received signal can be represented as:
\begin{equation}
	I+j\,Q = abs(I,Q) \cdot e^{j \times ang(I,Q)} 
\end{equation}

@@

The resulting values are sent through the bus to the decoder. After the
decoder received the samples it decides when a frame is sent or when the
channel is idle.  
The decoder also includes a function to correlate the
received bitstream with an expected pattern. Hence it is able to
determine the bit error ratio (BER). 
Furthermore the code we have
written so far subsumes different simulators for e.g. playing back
recoreded data. Another handy utility, which has been written in Octave,
is a tool (an oscilloscope) to look at the data, which is written into
files by the demodulator. An example of a recorded data stream with 25
kS/s can be seen in figure
\ref{fig:transmission}.  
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{./fig/transmission}
\caption{Start of a transmission of 101010.. with 1 kbit/s with idle state before. Data from the Octave oscilloscope. Amplitude is quite low with an average of 12 \% of the maximum, which is 127.5. This is due to a low signal strength. Sampling frequency is 25 kHz.}
\label{fig:transmission}
\end{figure}

@@

We have used the \textit{RTL-SDR} to observe the space variations of the
signal strength. \textit{Octave} scripts have been written to aid in
this purpose. These scripts are available online under
\url{https://github.com/s3xm3x/RTLSDRspecAn}. The scripts perform a
frequency sweep through the desired band. The obtained samples are then
transferred from the time domain to the frequency domain using a fast
fourier transformation (FFT).

Because the RTL-SDR has a maximum (stable) sampling rate of 2.4 MS/s, we are
limited to a maximum simultaneous acquisition band of 1.2 MHz. We keep a safety
margin to the maximum capabilities and sample the desired frequency band at 900
kHz intervals and stitch these bands together to form the overall frequency
sweep. This approach is valid only because we are not interested in real-time
data.  With this approach we are able to scan a 20 MHz band in roughly 30
seconds.

During the scanning process the gain is set to 40.2 dB. To get the
signal strength we simply take the average over a 10 MHz band around the
center frequency of the desired TV signal. This approach can be
justified by the fact, that DVB-T is specified up to 10 MHz bandwidth.
And the local TV signal provider is to its own statements using DVB-T 2.
The signal \ensuremath{|H|} is calculated as follows.  
\begin{equation}
	|H| = \Biggl| \sum_{n=0}^N FFT\biggl( \Re\{ s_{\text{band,n}}(t) \} \biggr) \Biggr|
\end{equation}     where \ensuremath{N} is the total number of intervals
accumulated over the entire frequency sweep. The FFT is only carried out over
the real values, which are received from the \textit{RTL-SDR}.
\ensuremath{s_{\text{band,n}}(t)} is the signal in the time domain of one
interval (one subband). There are as many samples taken in one band as needed
for an FFT of size 2048. To finally get \ensuremath{|H_{tv,avg}|} one has to
normalize everything and calculate the power. Before doing that a simple
smoothing algorithm (FIR) is applied on \ensuremath{|H|}.  
\begin{equation}
	|H_{tv,avg}| = 20 \cdot log (|H|) - 20 \cdot log(|H_0|)
\end{equation}   
where \ensuremath{|H_0|} is a normalization value measured at a certain point
in space. The result of combining measured signal strength values with the
haversine function to calculate the distance to the point where the maximum
signal has been captured can be observed in figure \ref{fig:haversine}.
\begin{figure}[h]
	\centering
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/haversine}
	\end{minipage}
	\hfill
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/heatmap_626mhz}
	\end{minipage}
	\caption{Left: Signal strength fading against distance relative
to the spot where the maximum signal strength has been captured (TV
towers). 474 MHz and 626 MHz belong to a tower in Uppsala, Vedyxa and
594 MHz belongs to a tower in Uppsala, Rickomberga. Distances have been
calculated with the Haversine formula. Right: Heatmap of the 626 MHz TV
signal in Uppsala.}
